A free/open-source Kazakh-Tatar machine translation system

### 2013-04-07 paper/paper.tex
some orthographic stuff

Tatar is spoken in and around Tatarstan by
approximately 6 million people. It is co-official with Russian in Tatarstan ---
a republic within Russia.  A majority of native speakers of both languages are bilingual in Russian.

@@

\subsection{Orthographic differences}
a generalised summary, 1 or 2 small specific examples
\subsection{Morphotactic differences}
2 or 3 specific examples
\subsection{Lexical and morphological differences}
2 or 3 specific examples
\subsection{Syntactic differences}
2 or 3 specific examples
> ұна- vs.\ ярат- (does this count as syntactic?)

>>>>

Tatar is spoken in and around Tatarstan by
approximately 6 million people. It is co-official with Russian in Tatarstan ---
a republic within Russia.  A majority of native speakers of both languages are bilingual in Russian. %, and the standard varieties our system FIXMEs are written in Cyrillic.

@@

\subsection{Orthographic differences}
%a generalised summary, 1 or 2 small specific examples
The standard varieties of Kazakh and Tatar our system deals with are both written in Cyrillic, though their implementations of Cyrillic differ in many ways.

While Tatar and Kazakh both have a velar/uvular obstruent distinction (e.g., /k/ vs.\ /q/) that interacts with adjacent vowels, the Tatar orthography only has one series of letters (e.g., ‹к›), relying on adjacent vowels (and employing ``hard'' and ``soft signs'' when these fail) to differentiate the two, and Kazakh has two series of obstruents (e.g., ‹к› and ‹қ›).

Kazakh does not orthographically distinguish high unrounded vowels (/ɘ/ ‹і› and /ə/ ‹ы›) before glides (/w/ ‹у› and /j/ ‹й›) by writing the combination with one letter; i.e., /ɘj/ and /əj/ are both written ‹и›, while /ɘw/ and /əw/ are both written ‹у›.  The quality of these vowels is necessary to know in order to predict the quality of following harmonising vowels, so this convention presents accute challenges to producing an accurate morphological transducer.

Tatar and Kazakh both use ``yoticised'' vowels---i.e., when ‹о›, ‹у›, or ‹а› (along with ‹ә› in Tatar) follow /j/, a single character is used to represent both: ‹ё›, ‹ю›, and ‹я› respectively.\footnote{Furthermore, in Tatar, /j/ followed by ‹э› or ‹ы› in Tatar is represented by ‹е›.}

\subsection{Morphotactic differences}
2 or 3 specific examples
\subsection{Lexical and morphological differences}
2 or 3 specific examples
\subsection{Syntactic differences}
2 or 3 specific examples
> ұна- vs.\ ярат- (does this count as syntactic?)
